Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-02 paper.tex

\subsection{The Li and Stephens model}
The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
coalescent with recombination~\citep{Hudson1983-properties}, which captures
many of the important features of the joint processes of mutation and
recombination. It is a Hidden Markov Model (HMM) in which a focal genome is a
sequence of nucleotides that are probabilistically emitted as
an imperfect mosaic of a set of reference genomes
(Figure~\ref{fig:ls_diagram}).
The LS model is used in a wide variety
of applications in genomics, including modern approaches to
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
and estimation of parameters such as
recombination rates \citep[e.g.][]{Hinch2011-tz},
and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.
See~\cite{Mcvean2019-linkage} for further review and discussion
of the LS model.

@@

@@

\subsection{Data preprocessing}
The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
\url{https://doi.org/10.55876/gis8.230329cd} (Supplementary Table 1).
We removed sequences
if they (1) had ambiguous collection dates;
(2) were collected before January 01, 2020 or after June 30, 2022; or
(3) were isolated from a non-human host.
We aligned sequences to the Wuhan-Hu-1/2019 reference sequence
(GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, we also excluded sequences if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').

% FIXME this isn't quite right
We encode ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
Problematic bases in the alignments, which had two or more ambiguous letters or
gaps within a distance of seven bases, were masked as missing data following
the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
were ignored when attaching a sample genome to the ARG, but they were
subsequently imputed. Further, 481 problematic sites flagged as prone to
sequencing errors or highly homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsection{Filtering ``time travellers''}
\label{sec:filtering_time_travellers}
[FIXME I'm pretty sure we don't have two filters here, we just use
maximum sampling delay. Also not sure the MSD acronym is worth it,
we don't use it much.]

Including samples with incorrectly reported collection dates (e.g., an Alpha
sample purportedly collected in 2020 from the United States, before Alpha
appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
artefacts that distorted the topology of ARGs inferred early on in the study.
Hence, to exclude such potential ``time travellers'' we employed two filters.
The first was to remove any sequence with a collection date that pre-dated the
time to the most recent common ancestor (tMRCA) of its corresponding clade. The
tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
(downloaded on August 22, 2022) where we used the lower bound of the 95\%
confidence interval of each clade as the minimum date cut-off. The second
filter was a simple threshold on the time delay between the collection date and
submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
inferred from the MSD-filtered genome set had noticeably fewer branches with a
high number of reversions. After some preliminary analysis, a MSD threshold
value of 30 days was used when building the ARGs described herein. Future work
could estimate collection dates by adding these potential time travellers back
on the tree, allowing an automated assessment of collection date discrepancies.

\subsection{Imputation of Pango lineage for non-sample nodes}
Initially, only sample nodes have Pango lineage assignments.
For analysis of recombinant parents, it is also helpful to assign
Pango lineages to non-sample nodes. We do this
using a two-step procedure (Figure~\ref{fig:imputation}). If the Pango
lineage of the parental node was available and no mutation was immediately
ancestral to the non-sample node, then the parental node’s status was directly
propagated to the non-sample node. Otherwise, the Pango lineage of the
non-sample node was inferred by matching its full set of mutations against the
Pango lineage-defining mutations (based on 90\% consensus of the sequences
analysed) from the COVID-CG website \citep{Chen2021-zc};
\url{https://covidcg.org/}; accessed on November 04, 2022).

\begin{figure} \centering
\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
Pango lineage for inserted, non-sample nodes. Here, three nodes (question
marks) have unknown Pango lineage (A). Because the parent of node 1 is
available (i.e., L2) and there is no mutation occurring on the edge connecting
the two nodes, the Pango lineage of the parent is directly propagated to node 1
(B). However, if there is a mutation (red X’s), then the Pango lineage is
inferred by comparison to the set of lineage-defining mutations from
COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
propagated to any non-mutant child nodes, as in case (D).
}
\end{figure}

>>>>

\subsection{The Li and Stephens model}
\label{sec:ls}
The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
coalescent with recombination~\citep{Hudson1983-properties}, which captures
many of the important features of the joint processes of mutation and
recombination. It is a Hidden Markov Model (HMM) in which a focal genome is a
sequence of nucleotides that are probabilistically emitted as
an imperfect mosaic of a set of reference genomes
(Figure~\ref{fig:ls_diagram}).
The LS model is used in a wide variety
of applications in genomics, including modern approaches to
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
and estimation of parameters such as
recombination rates \citep[e.g.][]{Hinch2011-tz},
and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.
See~\cite{Mcvean2019-linkage} for further review and discussion
of the LS model.

@@

[ALSO MENTION MISSING DATA, where we deal with Ns in the alignment]

@@

\subsection{Data preprocessing}
The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
\url{https://doi.org/10.55876/gis8.230329cd} (Supplementary Table 1).
We removed sequences
if they (1) had ambiguous collection dates;
(2) were collected before January 01, 2020 or after June 30, 2022; or
(3) were isolated from a non-human host.
We aligned sequences to the Wuhan-Hu-1/2019 reference sequence
(GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). We also excluded sequences if they had a
``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').

We encode ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments as missing data
(N). Problematic bases in the alignments, which had two or more Ns or
gaps within a distance of seven bases, are masked as missing data following
the approach used in the ``faToVCF`` tool used by
UShER~\citep{Turakhia2022-it}.
Sites that are masked by this process are
treated as missing data by the LS HMM (Section~\ref{sec:ls}).
In addition, we exclude 481 problematic sites flagged as prone to
sequencing errors or as highly homoplasic
(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2/},
accessed 2022-09-22) entirely.

Although the masking strategy used currently is simple and robust,
there are significant disadvantages, excluding, for example, any
deletions of length greater than one base. Exploring more sophisticated
masking strategies is an important route for future improvements.

\subsection{Filtering ``time travellers''}
\label{sec:filtering_time_travellers}
A major source of error in early versions of \texttt{sc2ts} was
the existance of ``time traveller'' sequences: those with erroneously
early collection dates.
For example, an Alpha sample purportedly collected in 2020 from the United States,
before Alpha appeared in the United Kingdom (USA/MN-Mayo-1563/2020)
produced significant topological distortions in
ARGs inferred early on in the study.
Hence, to exclude such potential ``time travellers'' we employ two filters.
The first is to remove any sequence with a collection date that pre-dates the
time to the most recent common ancestor (tMRCA) of its corresponding clade.
We obtained the tMRCA for each clade from a Nextstrain GISAID reference tree
(downloaded on August 22, 2022), and we used the lower bound of the 95\%
confidence interval of each clade as the minimum date cut-off.
This removed NN [Katherine can you fill in please?] sequences.

This filter removes the most egregious time travellers, but many still
remain. We thus employed a second filter which is a simple threshold on
the time delay between the collection date and submission date.
After some preliminary analysis we settled on a maximum submission delay
of 30 days when building the ARGs described here as a reasonable compromise.

As the results of \texttt{sc2ts} are sensitive to the existance of
time-travellers, an imporant aspect of future work is to find better ways
to identify them. One possibility is to use the LS HMM itself to flag
overly divergent sequences, and to exclude them from attachment to the ARG.
We might also estimate collection dates by adding these potential time travellers
back to the ARG, allowing an automated assessment of collection date discrepancies.

\subsection{Imputation of Pango lineage for non-sample nodes}
% FIXME the caption and the text are basically saying the same thing here. Can
% we streamline, and maybe give a bit more detail in the text about iteration,
% and the concrete implementation?
\begin{figure} \centering
\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
Pango lineage for inserted, non-sample nodes. Here, three nodes (question
marks) have unknown Pango lineage (A). Because the parent of node 1 is
available (i.e., L2) and there is no mutation occurring on the edge connecting
the two nodes, the Pango lineage of the parent is directly propagated to node 1
(B). However, if there is a mutation (red X’s), then the Pango lineage is
inferred by comparison to the set of lineage-defining mutations from
COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
propagated to any non-mutant child nodes, as in case (D).
}
\end{figure}
For many applications it is helpful to infer the Pango lineage status
of the internal ARG nodes estimated by \texttt{sc2ts}.
% Initially, only sample nodes have Pango lineage assignments.
% For analysis of recombinant parents, it is also helpful to assign
% Pango lineages to non-sample nodes.
We do this using a two-step procedure (Figure~\ref{fig:imputation}).
Initially, only sample nodes have Pango lineage assignments.
For a given non-sample node $u$, if the Pango lineage of its parent is
available and there are no intervening mutations,
then we impute that $u$'s Pango lineage is the same as its parent.
Otherwise, the Pango lineage of the
non-sample node is inferred by matching its full set of mutations against the
Pango lineage-defining mutations (based on 90\% consensus of the sequences
analysed) from the COVID-CG website \citep{Chen2021-zc};
\url{https://covidcg.org/}; accessed on November 04, 2022).

[TODO how accurate is this approach? We should probably discuss it somewhere
in the Results, and refer back from here? ]
